\chapter{Ethical considerations}
\label{chap:ethics}

Overall, there does not seem to be any major ethical concerns regarding the nature of 
this project. We have also investigated the licenses for the libaries we intend to incorporate
within our tool, namely Mythril and the Solidity compiler, which respectively have the MIT and GPL-3.0
licenses, and should be permissive enough for our nature of work. However, there are still certain key points of consideration. First of all,
because the project deals with decompiling Solidity code into its Yul intermediate
representation and EVM code, and generating the mappings between them, they may reveal or
expose vulnerabilities within the compiler generated code. With Ethereum smart contracts 
nowadays holding up to millions of dollars worth of funds, if such information is revealed
to malicious actors, it could result in a permanent and irrevocable loss of funds.

In addition, this feature may also allow malicious actors to more easily obfuscate their code,
by possibly implanting fragments of malicious inline assembly into template contract codes,
claiming they serve another purpose. Then, users who do not analyse the code in detail and
simply reuse them may at best lead to bugs in execution, or at worst loss of funds.

Next, there is no guarantee provided by our project about the precision or reliability
of the gas analysis generated. Therefore, if the gas analysis is not accurate, and other projects
do not carefully consider the gas used by their contract calls and simply rely on estimates derived
from our tools, it may result in out of gas errors in future transactions, which may in turn
lead to funds being locked forever and lost. We therefore intend to waive any liabilities derived
from the use of our tool.

Also, generating the Yul intermediate representation is still an experimental function of
the Solidity compiler that we make use of. Therefore, it is not assured to be completely
correct, and users should not use the bytecode generated by our code directly when
deploying contracts.

Finally, when collecting user data during the evaluation of our project, we must also consider
the potential privacy risks and the right to withdraw from the survey. Users will not be asked to 
provide any personal information, and the data collected will be anonymised and aggregated, to
prevent any leaks of sensitive information.